---
title: "LeggettPlus: A Guide to the Lead PBPK Simulator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LeggettPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`LeggettPlus` provides an open-source, reproducible implementation of the Leggett+ physiologically based pharmacokinetic (PBPK) model for lead (Pb). The model was originally developed by [Richard Leggett (1993)](https://pmc.ncbi.nlm.nih.gov/articles/PMC1519877/) and later updated by [Vork, Carlisle and Brown (2013)](https://oehha.ca.gov/sites/default/files/media/downloads/air/document/pbpk2013.pdf); [Vork & Carlisle (2020)](https://doi.org/10.1080/15459624.2020.1743845); and [Vork, Brown, and Carlisle (2023)](https://doi.org/10.1080/15459624.2022.2150767).

The package includes:

-   `leggett_base_parameters()` — build the physiological parameter set
-   `run_leggett_model()` — solve the 21-compartment ODE system across one or more exposure stages
-   `summarize_leggett_output()` — extract compartment values, maxima, and AUC from a model run
-   `launch_leggett_app()` — start the bundled Shiny application

## Intended use and limitations

LeggettPlus is intended for research, regulatory evaluation, and exposure-scenario exploration. Outputs should be interpreted within the context of model assumptions, parameter uncertainty, and available toxicokinetic data. The package is **not** intended for clinical diagnosis or individual medical decision-making.

**Male-reference physiology.** The package's physiological parameters and calibration constants (body weight, hematocrit, blood-volume scaling, and RBC binding capacity) are derived from adult male / male-dominated occupational-worker data (Leggett, 1993, and subsequent updates). There is no female-specific parameter set implemented — no separate hematocrit range, blood-volume expansion, or RBC binding capacity for female individuals, including during pregnancy. Simulated results should not be assumed to represent female physiology.

------------------------------------------------------------------------

# Model structure: 21 compartments

The ODE system tracks lead mass (µg) through the following compartments:

| Code | Compartment | Biological meaning |
|----------------|----------------------|----------------------------------|
| `plasd` | Plasma (diffusible) | Free, unbound lead in plasma; primary distribution pool |
| `plasb` | Plasma (bound) | Protein-bound lead in plasma |
| `evf` | Extracellular fluid | Interstitial fluid exchanging rapidly with plasma |
| `rbc` | Red blood cells | Lead sequestered in erythrocytes; major blood burden |
| `csf` | Cortical bone surface | Rapid surface exchange with plasma |
| `cex` | Cortical bone (exchangeable) | Intermediate-term cortical storage |
| `cne` | Cortical bone (non-exchangeable) | Long-term cortical retention (years–decades) |
| `tsf` | Trabecular bone surface | Rapid surface exchange with plasma |
| `tex` | Trabecular bone (exchangeable) | Intermediate-term trabecular storage |
| `tne` | Trabecular bone (non-exchangeable) | Long-term trabecular retention |
| `livhu` | Liver (high perfusion) | High-turnover hepatic pool |
| `livlu` | Liver (low perfusion) | Slow-turnover hepatic pool |
| `strpd` | Soft tissue (rapid) | Fast-exchange peripheral tissues |
| `stmid` | Soft tissue (intermediate) | Medium-turnover soft tissues |
| `stslo` | Soft tissue (slow) | Slow-exchange depot (e.g., hair, nails) |
| `brain` | Brain | Central nervous system compartment |
| `kidoth` | Kidney (other) | Non-excretory renal tissue |
| `kidurp` | Kidney (urinary path) | Proximal tubule + urinary path excretion |
| `sint` | Small intestine | GI absorption zone; receives oral + biliary load |
| `lintu` | Large intestine (upper) | Transit toward excretion |
| `lintl` | Large intestine (lower) | Final GI compartment; lead excreted in feces |

Blood lead level (BLL, µg/dL) is a derived quantity computed from `rbc + plasd + plasb` divided by blood volume. It is not a state variable but is calculated during summarization.

------------------------------------------------------------------------

# Quick start

## Minimal simulation

A simulation requires a **stage schedule** — a data frame where each row represents a discrete exposure period (stage). All columns must be present.

```{r}
library(LeggettPlus)

# Single-stage example: background oral intake only, no inhalation
schedule <- data.frame(
  # ---- Time ----
  duration = 365,               # Length of this stage in DAYS (integer; rounded if fractional)

  # ---- Non-inhalation background intakes (µg/day) ----
  # These represent all non-air occupational sources: ingestion of
  # contaminated dust, dermal absorption, dietary lead at the worksite, etc.
  Occ.background.ug.d = 0,      # Occupational background (µg/day) — set 0 if air is the only source

  # Non-occupational background: diet, drinking water, consumer products, soil/dust
  # ingestion at home. FDA interim reference level for adults is 8.8 µg/day from food.
  NonOcc.background.ug.d = 2.2, # Non-occupational background (µg/day)

  # ---- Airborne lead concentrations (µg/m³) ----
  Occ.pb.ug.m3    = 0,          # Airborne Pb at the workplace (µg/m³);
                                 # OSHA PEL = 50 µg/m³; OSHA action level = 30 µg/m³

  NonOcc.pb.ug.m3 = 0.15,       # Ambient airborne Pb outside work (µg/m³);
                                 # NAAQS rolling 3-month average = 0.15 µg/m³

  # ---- Breathing rates (m³/DAY) ----
  # The model uses daily rates. Convert hourly rates by multiplying by 24.
  # OEHHA Air Toxics Hot Spots Guidance (2015) provides age/activity-specific
  # rates in L/kg-day (residential) or L/kg-8h (worker) that must first be
  # converted to m³/day using body weight.
  Occ.breath.rate.m3.d    = 30, # Occupational (worker) daily breathing rate (m³/day)
  NonOcc.breath.rate.m3.d = 17, # Non-occupational (residential) daily breathing rate (m³/day)

  # ---- Work schedule ----
  work.hour.perweek = 40,        # Hours worked per week; used to partition
                                 # occupational vs. non-occupational exposure time.
                                 # Set to 0 for scenarios without a workplace.

  # ---- Oral intake (µg/day) ----
  # Direct oral uptake entering the GI tract. Absorb fraction is controlled by
  # abs_ratio in run_leggett_model() (default 0.20 for adults).
  oral.intake.ug.d = 0           # Any additional direct oral dose not already in
                                 # the background columns (e.g., a supplement or
                                 # experimental bolus). Usually 0 for field scenarios.
)

out <- run_leggett_model(schedule)
head(out$time_series)
```

### What `run_leggett_model()` returns

The function returns a named list:

-   **`time_series`**: a data frame with one row per day and one column per compartment (plus `time` in days). Values are lead mass in **µg**.
-   **`metadata`**: model inputs and derived volumes (`blood_volume_dl`, `plasma_volume_dl`, `rbc_volume_dl`) needed for concentration calculations.

------------------------------------------------------------------------

# Physiological parameters

`leggett_base_parameters()` computes static physiological values and the initial state vector. You can inspect or override defaults, but note that `body_weight_kg` and `hematocrit` are both **male-reference values**: the defaults below, and the RBC binding saturation constant they feed into (`Sat.ug.dL` in `leggett_base_parameters()`), are calibrated against adult male / occupational-worker data. There is no female-specific hematocrit range or blood-volume expansion implemented.

```{r}
params <- leggett_base_parameters(
  body_weight_kg    = 73,     # Body weight (kg). Scales blood volume and OEHHA breathing-rate conversions.
                              # Reference value from adult male occupational-worker data.
  hematocrit        = 0.3802, # Fractional hematocrit (0–1). Determines RBC vs. plasma volume split.
                              # Default is the reference adult male value used in Leggett (1993);
                              # no female-specific default is provided.
  age_years         = 40,     # Age at start of simulation (years). Influences some kinetic parameters.
  initial_bll_ug_dl = 0.78    # Initial blood lead level (µg/dL) before new exposures begin.
                              # CDC NHANES 50th-percentile for US adults (2015–16) is 0.78 µg/dL.
)

# Derived blood volumes (dL)
params$blood_volume_dl
params$rbc_volume_dl
params$plasma_volume_dl
```

Additional parameters passed to `run_leggett_model()`:

| Argument | Default | Meaning |
|--------------------------|-----------------------|-----------------------|
| `inhalation_transfer` | 0.30 | Inhalation Transfer Coefficient (ITC): fraction of inhaled lead absorbed into plasma. Calibrated to \~0.30 from occupational worker data. |
| `abs_ratio` | 0.20 | Oral absorption fraction for GI uptake. Adults typically absorb \~20% of ingested lead. |

------------------------------------------------------------------------

# Multi-stage schedules

Each row in the schedule data frame is an independent exposure stage solved sequentially. The final compartment state of one stage becomes the initial state for the next. This allows you to model changing exposures over a career lifetime.

## Occupational scenario

An example occupational scenario is provided below: 20 years of workplace exposure followed by 20 years of retirement with background exposure only.

```{r}
schedule_occ_ret <- data.frame(
  duration = c(
    20 * 365,  # Working years (days)
    20 * 365   # Retirement years (days)
  ),
  Occ.background.ug.d = c(25, 0),    # Occupational non-air background: 25 µg/day at work, 0 in retirement
  NonOcc.background.ug.d = c(2.2, 2.2), # Non-occupational background stays constant
  Occ.pb.ug.m3 = c(30, 0),           # Occupational airborne Pb: 30 µg/m³ during work, 0 in retirement
  NonOcc.pb.ug.m3 = c(0.15, 0.15),   # Ambient Pb stays constant
  Occ.breath.rate.m3.d = c(30, 0),   # Worker breathing rate during exposure; 0 when retired
  NonOcc.breath.rate.m3.d = c(17, 17),
  work.hour.perweek = c(40, 0),       # 40 h/week during work; 0 in retirement
  oral.intake.ug.d = c(0, 0)
)

out_ret <- run_leggett_model(
  stage_schedule    = schedule_occ_ret,
  body_weight_kg    = 73,
  hematocrit        = 0.3802,
  age_years         = 30,             # Age at simulation start (will be 30 when work begins)
  initial_bll_ug_dl = 0.78
)

nrow(out_ret$time_series)  # total days simulated
```

## Pulse scenario

A pulse scenario models discrete, acute exposure events separated by recovery intervals. Each pulse is represented as a one-day stage with an elevated `Occ.background.ug.d` equal to the pulse dose.

```{r}
# Background stage between pulses
background_stage <- data.frame(
  duration                = 29,   # days between pulses
  Occ.background.ug.d     = 0,
  NonOcc.background.ug.d  = 2.2,
  Occ.pb.ug.m3            = 0,
  NonOcc.pb.ug.m3         = 0.15,
  Occ.breath.rate.m3.d    = 0,
  NonOcc.breath.rate.m3.d = 17,
  work.hour.perweek       = 0,
  oral.intake.ug.d        = 0
)

# Pulse stage: same as background but with a one-day dose
pulse_stage <- background_stage
pulse_stage$duration <- 1
pulse_stage$Occ.background.ug.d <- 500  # µg delivered in one day

# 3 pulses each separated by 29 background days
pulse_schedule <- do.call(rbind, replicate(3, rbind(pulse_stage, background_stage), simplify = FALSE))

out_pulse <- run_leggett_model(pulse_schedule)
```

------------------------------------------------------------------------

# Breathing rates

The model requires **daily** breathing rates (m³/day) for occupational and non-occupational periods. The OEHHA Air Toxics Hot Spots Guidance (2015) provides age- and activity-specific rates that can be converted as follows:

**Residential (non-occupational):** Table 5.6 gives rates in L/kg-day. Convert to m³/day:

```{r eval=FALSE}
# L/kg-day × body_weight_kg / 1000 = m³/day
# Example: 16–70 y mean = 185 L/kg-day, 73 kg body weight
residential_m3_d <- 185 * 73 / 1000
```

**Worker (occupational):** Table 5.8 gives 8-hour rates in L/kg-8h. Convert to m³/day:

```{r eval=FALSE}
# L/kg-8h × body_weight_kg / 8 h × 24 h/day / 1000 = m³/day
# Example: Moderate intensity (3.0–<6.0 MET), 95th = 230 L/kg-8h, 73 kg
worker_m3_d <- 230 * 73 / 8 * 24 / 1000
```

The Shiny app automates this conversion when **OEHHA mode** is selected, using the appropriate age band and activity intensity from Tables 5.6 and 5.8. Job presets map common occupational categories (e.g., Farming, Office Work) to Table 5.8 intensity bands.

------------------------------------------------------------------------

# Worker Adjustment Factor and Discount Factor (WAF/DF)

When a lead source operates intermittently (e.g., 8 h/day, 5 days/week), the annual average concentration over-represents exposure. OEHHA Eq. 5.4.1.2 defines the **Worker Adjustment Factor (WAF)**:

```         
WAF = (24 / H_source) × (7 / D_source) × DF
```

where:

-   `H_source` = hours per day the source is active
-   `D_source` = days per week the source is active
-   `DF` = Discount Factor (1.0 unless partial shift overlap applies)

For a standard 8 h/day, 5 d/week scenario with full overlap, WAF ≈ 4.2. The app's WAF/DF block in the Exposure Inputs panel computes this automatically and adjusts the occupational air concentration accordingly.

------------------------------------------------------------------------

# Summarizing results

`summarize_leggett_output()` extracts a table of compartment statistics at a user-specified time point, plus maxima and AUC across the run.

```{r}
# Extract summary at day 365
summary_df <- summarize_leggett_output(
  time_series      = out$time_series,
  time_point       = 365,              # Day at which to report compartment values
  blood_volume_dl  = out$metadata$blood_volume_dl,   # Required to compute BLL (µg/dL)
  plasma_volume_dl = out$metadata$plasma_volume_dl,  # Required to compute plasma concentration
  rbc_volume_dl    = out$metadata$rbc_volume_dl      # Required to compute RBC concentration
)

# Derived quantities appended to the table:
#   "Blood Lead (ug/dL)"       — (rbc + plasd + plasb) / blood_volume_dl
#   "Red Blood Cell Lead (ug/dL)" — rbc / rbc_volume_dl
#   "Plasma Lead (ug/dL)"     — (plasd + plasb) / plasma_volume_dl
summary_df
```

Columns returned:

| Column | Meaning |
|----|----|
| `Compartment` | Human-readable compartment name |
| `Value` | Lead mass (µg) or concentration (µg/dL) at `time_point` |
| `Max` | Maximum value across the entire simulation |
| `Time_of_Max_days` | Day on which the maximum occurred |
| `AUC` | Trapezoidal area under the time–mass curve (µg·days) |

------------------------------------------------------------------------

# Choosing a scenario type

Three scenario templates are available (in the app or via the schedule data frame directly):

**Occupational** — Steady workplace and non-occupational exposures over a fixed duration. Use this when you have a single continuous exposure period with no planned change. Key inputs: airborne concentrations, breathing rates, background intakes, and working hours per week.

**Occupational with Retirement** — Adds a post-work stage where occupational air and background are set to zero (or to a lower ambient level). Use this to project long-term BLL decline after cessation and to assess whether bone mobilization keeps BLL elevated. Key decision: how many retirement years to simulate, and whether to retain ambient non-occupational exposure.

**Pulse** — Models discrete, time-limited exposure events separated by recovery intervals. Use this for sporadic high-dose situations (e.g., occasional welding, remediation activities, or experimental bolus doses). Key decisions: pulse dose (µg), number of pulses, and inter-pulse interval (days). The background stage between pulses controls the baseline load.

------------------------------------------------------------------------

# Shiny application

The bundled Shiny app provides a graphical interface to all of the above, including:

-   Template-driven scenario builder (Occupational / Retirement / Pulse)
-   OEHHA breathing-rate lookup with job presets
-   WAF/DF adjustment panel
-   Interactive time-series plot (Plotly) with compartment group filters
-   Summary data table with downloadable CSV

Launch the app with:

```{r eval=FALSE}
launch_leggett_app()
```

The app requires the optional `shiny`, `bslib`, `plotly`, `DT`, `shinycssloaders`, and `glue` packages. Install them with:

```{r eval=FALSE}
install.packages(c("shiny", "bslib", "plotly", "DT", "shinycssloaders", "glue"))
```

------------------------------------------------------------------------

# References

Leggett, R.W. (1993). An age-specific kinetic model of lead metabolism in humans. *Environmental Health Perspectives*, 101(7), 598–616.

California OEHHA (2015). *Air Toxics Hot Spots Program Guidance Manual for Preparation of Health Risk Assessments*. Office of Environmental Health Hazard Assessment.